---
title: "The farmers against the fjord: cross-scale interactions of environmental gradients and management on grassland assemblages in small-scale grazing systems"
format:
  html:
    embed-resources: true
author:
  - name: Morgane Kerdoncuff
    orcid:
    email: morgane.kerdoncuf@uib.no
    affiliations:
      - Department of Biological Sciences, University of Bergen, Norway
      - Center for Sustainable Area Management (CeSAM), University of Bergen, Norway
  - name: Amy Elizabeth Eycott
    orcid:
    affiliations: 
      - Faculty of Biosciences and Aquaculture, Nord University, Norway
      - Center for Sustainable Area Management (CeSAM), University of Bergen, Norway
  - name: Adam Klimes
    orcid:
    affiliations: Department of Biological Sciences, University of Bergen, Norway
  - name: Inger Elisabeth Måren
    orcid:
    affiliations:
      - University of Bergen, Norway
      - UNESCO Chair on Sustainable Heritage and Environmental Management, University of Bergen, Norway
      - Center for Sustainable Area Management (CeSAM), University of Bergen, Norway
date: today
bibliography: NBR_survey.bib
execute: 
  echo: false
---

```{r ScriptSource}
#| include: false
source("scripts/NBR_TradMod_FjordSysCANCOR.R", local = knitr::knit_global())
```

```{r setup, include=FALSE}

# Register an inline hook:
knitr::knit_hooks$set(inline = function(x) {
  x <- sprintf("%1.2f", x)
  paste(x, collapse = ", ")
})

# Define a generic
custom_print <- function(x, ...) {
  UseMethod("custom_print", x)
}

# Define a method for p-values
custom_print.p.value <- function(x, ...) paste(sprintf("%1.3f", x), collapse = ", ")
# Define a method for t-statistics
custom_print.stat <- function(x, ...) paste(sprintf("%1.2f", x), collapse = ", ")

```

# ABSTRACT

Small-scale farming across the world have created agro-ecological landscapes of high ecological and cultural values that support local food production and sustainable land use. In Western Norway, farmers have adapted to the sharp climate-landform gradient of the fjord system essentially by relying on low-intensive rotational grazing of sheep, cattle and/or goats. If these management practices were shown to favour field biodiversity and productivity in controlled experiments, it is uncertain that their effects would be as straightforward in a context of high abiotic heterogeneity. We therefore present an exploratory study of the cross-scale effect of fjord climate, landscape composition, grazing management and fine-scale environment on the composition of 29 grazing fields of the Nordhordland Biosphere Reserve. We used canonical correlation and redundancy analyses to investigate relationships between the different explanatory sets and the dominant assemblages of grasses, forbs and beetles. We found that landscape composition around grazing fields was correlated with fjord climate and that the dominance of common plant species was affected by the inherent productivity and the edaphic conditions of the grazing fields. However, results regarding the influence of climate, landscape or grazing on plant and beetle assemblages were contradictory, highlighting the challenge of capturing complex cross-scale interactions with a simplified study design.

# INTRODUCTION

<!-- At the European scale, even though the Green Revolution did substantially increase food security over the second half of the XX^th^ century, the current degraded state of agricultural lands is rather forecasting stagnation if not decline of productivity potential for the coming decades [@jongmanHomogenisationFragmentationEuropean2002; @haladaWhichHabitatsEuropean2011; @vosTrendsEuropeanCultural1999]. Less emphasized is the impact of the agro-industrial model and trade on the people standing at the frontline of the chain, namely the farmers. While they should be amongst the most influential apparatus regarding land use and natural resource management, they find themselves at the bottom of the decision-making process, subdued to the versatility of the market and competing with fellow farmers from the other side of the globe [@holt-gimenezAgroecologyFoodSovereignty2012]. -->

<!-- First biosphere reserve in Norway, the Nordhordland UNESCO Biosphere Reserve (NBR) captures the singularity and variety of small-scale grazing systems in Western Norway [@kalandNordhordlandBiosphereReserve2018]. To adapt to the complex topographic and climatic gradients driven by the fjord geomorphology [@ketzlerClimateNorway2021], former farmer-fishermen communities relied on small, self-sustained farms, encompassing most of necessary resources and land use types for subsistence of the household [@hetlandPluriactivityStrategyEmployment1986]. Thus, typical farm units would encompass a few arable lands, some forested areas, more productive infields for hay-making and less productive outfields for livestock grazing, resulting in a both temporally and spatially dynamic mosaic-like landscape [@haugeGaldaneLaerdalWestern1988; @hjelleHesitantHuntersReview2006; @indrelidFarmingSystemIts1988; @olssonAgroecosystemsNeolithicTime1991]. Following the global productivist strategy promoted in the 1970s, the intensification of agricultural practices coupled with the socio-economic decline of rural areas initiated two antagonist trends, still ongoing nowadays [@bjorkhaugMultifunctionalAgriculturePolicy2008]. On one hand, infields have been enlarged and improved with the introduction of artificial fertiliser to support the shift towards market-oriented livestock production, while more remote outfields have been progressively abandoned, leading to a drastic decline of the number of holdings [@fjellstadPatternsChangeTwo1999; @forbordDriversChangeNorwegian2014; @hamreLandcoverStructuralChanges2007a; @jonesLandtenureLandscapeChange1988]. However, compared to the rest of Norway and other European countries, in Nordhordland agricultural intensification has remained limited due the environmental constraints, thus preserving at least to a better extent small-scale grazing practices and the highly valued fjord landscape [@dramstadTenancyNorwegianAgriculture2010; @fjellstadPatternsChangeTwo1999]. -->

Industrialised agriculture reliant upon globalized food chains has pushed the capacity of the Earth system beyond its limits [@steffenPlanetaryBoundariesGuiding2015; @ipbesGlobalAssessmentReport2019; @richardsonEarthSixNine2023]. The literature pool assessing its global impact on biodiversity [@beckmannConventionalLandUse2019; @kleijnRelationshipFarmlandBiodiversity2009; @newboldGlobalEffectsLand2015], with subsequent erosion of ecosystem functions and services [@jeanneretIncreaseFoodProduction2021; @flynnLossFunctionalDiversity2009], has been growing relentlessly over the past 20 years and has made fertile ground for alarming predictions if business-as-usual was to continue over the rest of this century [@egliWinnersLosersNational2018; @foleySolutionsCultivatedPlanet2011; @hunterAgriculture2050Recalibrating2017; @salaGlobalBiodiversityScenarios2000]. Following the repeated calls for a sustainable transition of agriculture in order to transcend short-term productivist strategies, a regained interest has grown towards small-scale farming systems [@cunninghamCloseYieldgapSaving2013; @garibaldiFarmingApproachesGreater2017; @kremenDiversifiedFarmingSystems2012; @torokGrazingEuropeanOpen2016]. By prioritising long term preservation of natural resources over short-term yield maximisation, these approaches enhance the heterogeneity and resilience of agro-ecosystems [@gardnerAgriculturalResiliencePerspectives2019; @menconiRevisitingGlobalFood2022], thus empowering local communities and supporting biodiversity of high value [@altieriLinkingEcologistsTraditional2004; @bignalLowintensityFarmingSystems1996; @sonninoLocalFoodscapesPlace2013]. By nature adaptative and tailored to the local environment, small-scale farming is not suited for standardisation and therefore requires an extensive effort of research across ecosystems to comprehend the whole spectrum of practices and how they interact with environmental dynamics and feedback [@altieriLinkingEcologistsTraditional2004; @cunninghamCloseYieldgapSaving2013; @gomesHeuristicsFarmHeterogeneity2022; @joseEcologicalConsiderationsSustainable2019]. A well-suited framework for this task is the UNESCO World Network of Biosphere Reserves (WNBR), which, through the Man and Biosphere (MAB) program, offers a platform for international collaboration and knowledge-sharing [@unescoNewRoadmapMan2017]. As model areas for regional experimentation and implementation of sustainable development [@barracloughGlobalKnowledgeAction2023; @unitednationsTransformingOurWorld2015], Biosphere Reserves promote applied research directed towards food security and locally adapted agricultural production [@anderzenEffectsOnfarmDiversification2020; @kratzerRuralInnovationsBiosphere2019; @tamouPastoralistsChangingEnvironment2018].

In Norway, the Nordhordland UNESCO Biosphere Reserve (NBR) showcases in a nutshell the diversity and singularity of small-scale grazing systems in westerly fjords [@kalandNordhordlandBiosphereReserve2018]. Strong climatic and topographic gradients first constrained farmer-fishermen communities to develop pluriactivity in small, self-sustained farms [@haugeGaldaneLaerdalWestern1988; @hetlandPluriactivityStrategyEmployment1986; @hjelleHesitantHuntersReview2006; @mehlCulturalLandscapeDevelopment2015a], before limiting agriculture intensification during the post-WW2 productivist era [@dramstadTenancyNorwegianAgriculture2010; @fjellstadPatternsChangeTwo1999]. The resulting agro-ecological landscape encompasses amongst other grazed grasslands, arable lands, hay-meadows and forested areas in a fine-grained mosaic holding both high ecological and cultural values [@cusensParticipatoryMappingReveals2022; @wehnAdaptiveBiodiversityManagement2018]. Small-scale grazing in semi-natural grasslands involves rotational management with low stocking density and high diversity of locally-adapted breeds of cattle, sheep and goats. These different practices, documented in both grazing experiments and observational studies, have shown not only to enhance biodiversity [@bakkerDoesResourceAvailability2003; @skornikEffectGrazingPressure2010; @wangDiversifyingLivestockPromotes2019], but also to increase field productivity [@boavistaEffectsGrazingRegimes2019; @skornikEffectGrazingPressure2010].

There is however a limit when assessing effect of grazing management independently from environmental context [@duelliBiodiversityEvaluationAgricultural1997] as species distribution in a given habitat results from intricate organism-environment interactions across different spatial and temporal scales [@freestoneRegionalEnrichmentLocal2006; @joseEcologicalConsiderationsSustainable2019; @oesterheldImpactGrazingSpecies2011]. For instance, as climatic gradients shape regional species pools [@blonderLinkingEnvironmentalFiltering2015; @habelEuropeanGrasslandEcosystems2013] and habitat productivity [@gaoChangesGlobalGrassland2016; @guoSpatialVariationsAboveground2012], factors such as geographical location [@birkhoferGeneralRelationshipsAbiotic2012], elevation [@austrheimPlantDiversityPatterns2002; @turtureanuScaleTaxondependentBiodiversity2014], temperature and precipitation [@dauberLocalVsLandscape2005; @wehnAdaptiveBiodiversityManagement2018; @zelnikPlantSpeciesDiversity2013] are likely to influence species distribution not only across the NBR, but also at finer scale through site specific conditions [@fynnManagingBrowsingGrazing2019; @reitaluDeterminantsFinescalePlant2014]. Likewise, surrounding landscape composition [@steffan-dewenterScaledependentEffectsLandscape2002; @leprovostContrastingResponsesBelowground2021; @soderstromPlantsInsectsBirds2001; @zulkaSpeciesRichnessDry2014] and structure [@billeterIndicatorsBiodiversityAgricultural2007; @concepcionEffectsLandscapeComplexity2008; @gonthierBiodiversityConservationAgriculture2014; @hassLandscapeConfigurationalHeterogeneity2018] not only affect the flow of grassland species but can also enrich the assemblage with specialists from other habitats. Lastly, in-field environmental conditions, either completely inherent (e.g. topography) [@klimekPlantSpeciesRichness2007; @turtureanuScaleTaxondependentBiodiversity2014] or affected by broader climatic gradients and grazing disturbance for instance (e.g. soil properties) [@dorroughPlantNeighbourhoodLandscape2007; @mayelHowDoesGrassland2021; @zelnikPlantSpeciesDiversity2013], are likely to shape stress levels and resource availability to organisms.

Thus, considering that small-scale agro-ecological systems are embedded within complex contexts of cross-scale influence and feedback between organisms, environmental gradients and human management, to what extent can Nordhordland farmers truly affect their grazing fields [@hustonLocalProcessesRegional1999; @siefertScaleDependenceVegetationenvironment2012]? When it comes to grazing systems, most studies have taken a conservation perspective within contexts of either exceptionally high biodiversity [@denglerBiodiversityPalaearcticGrasslands2014; @habelEuropeanGrasslandEcosystems2013; @turtureanuScaleTaxondependentBiodiversity2014] or agricultural intensification [@billeterIndicatorsBiodiversityAgricultural2007; @gonthierBiodiversityConservationAgriculture2014; @zulkaSpeciesRichnessDry2014], thus rather focusing on diversity and/or rarity indicators. The variation of ordinary plant and animal communities have been less thoroughly investigated [but see @kleinebeckerEffectsGrazingSeasonal2011; @klimekPlantSpeciesRichness2007; @zelnikPlantSpeciesDiversity2013] even though it offers both fundamental and applied prospects not only for habitat conservation [@jiguetInferencesCommonSpecies2006] but also towards the ecosystem function and service frameworks [@avolioChangesPlantCommunity2014; @phoenixSustainingEcosystemServices2008]. Indeed, from the farmer's point of view the most important features of grazing fields are the quality and quantity of fodder, which largely depends on plant assemblage [@collombCorrelationFattyAcids2002; @frenchSpeciesCompositionDetermines2017; @marrsReleaseSheepGrazing2020] and available resources such as soil nutrients, water or light availability [@avolioChangesPlantCommunity2014; @bakkerDoesResourceAvailability2003; @fayGrasslandProductivityLimited2015; @radujkovicSoilPropertiesKey2021]. Understanding what influences the fields functionality and productivity would undoubtedly benefit the preservation of small-scale grazing systems, which mainly relies on farmers' ability and willingness to pursue their activity [@mcginlayThreatAbandonmentSocioecological2017; @sinclairOptionsContextApproach2019; @stensekeBiodiversityLocalContext2006].

In this study we attempted to draw a preliminary picture of the different factors influencing dominant grass, forb and beetle assemblages of grazing fields in the NBR. These taxa were chosen according to their contributions to forage production, nitrogen capture and/or pollination, and nutrient recycling through dung decomposition respectively. We categorised environment and management variables into sets according to four spatial scales (@fig-fjordconcept): the **fjord system**, which encompasses climatic and landform features operating at the regional scale; the **landscape matrix**, which considers land uses around the farm; the **grazing management**, related to livestock breeding at the field and farm scales; and finally the **fine-scale environment** which includes edaphic, topographic and ground cover parameters at the level of the sampling area. Our main focus was to question whether grazing management would hold greater influence on plant and beetle assemblages compared with underlying environmental gradients, and if grazing management itself would be affected by them.

![Conceptual model outlining the cascading influence of the fjord, landscape, grazing and fine-scale environment gradients on dominant grass and forb species and dominant beetle families of grazing fields in the NBR fjord system. All variables prior to selection are listed, with black ones included and grey ones excluded from the analysis after selection. For description and selection of explanatory variables and response assemblages, see the Material and Methods section. Credits: figure created on Inkscape v. 1.1 (Inkscape Developers, https://inkscape.org)](illustrations/NBR_FjordConcept.png){#fig-fjordconcept}


# MATERIAL AND METHODS

## Study area

The Nordhordland UNESCO Biosphere Reserve (NBR) is located in the fjord system of Western Norway, enclosed between the Bergen region in the south and the Sognefjorden, the deepest Norwegian fjord, in the north [@kalandNordhordlandBiosphereReserve2018]. There are sharp elevation and climatic gradients from the flat coastal plateau on the west to the eastern mountains, with decreasing average July temperature, ranging from 15.9 to 8.3°C, and increasing annual precipitations, ranging from 1550 to 4800 mm [@ketzlerClimateNorway2021; @meteorologiskinstituttNorskKlimaservicesenter2023]. Our study area is located in the lowlands of the southern part of the NBR, under 250 masl.

In summer 2019 and 2020, we conducted an ecological survey on 29 grazing fields across the fjord system of the NBR (@fig-sitemap). Our sites were defined as "infields" (*innmarksbeite*), which are "*agricultural land that cannot be harvested mechanically* \[with\] *at least 50% of the area* \[...\] *covered with cultured grass or pasture-tolerant herbs*" according to the Norwegian "areal type" AR5 classification system [@ahlstromAR5Klassifikasjonssystem2019]. All fields were grazed for at least 10 years prior to sampling, with the same livestock type, either sheep, goats or cattle, for at least the last 5 years. According to the farmers, most fields had been continuously grazed for at least 50 years; only three had a short grazing interruption (5 years or under) and one had a long grazing interruption (20 years). Grazing intensity was low, with a maximum stocking rate of 0.6 yearly unit per ha. Twenty-one sites were enriched at least occasionally with single or simultaneous use of artificial fertiliser, manure and shellsand-lime mix. To the knowledge of the farmers, no field had been resown and/or plowed since at least the 1970s, and only four out of the 29 sites received upgraded drainage between 1990 and 2010. More details on the management and the different treatments applied to the fields are available in the project data paper [**ref**].

![Study sites within Nordhordland UNESCO Biosphere Reserve, delimited in white, in Western Norway. Colour codes are for livestock types: black for cattle, grey for sheep, white for goats. Credits: map created in QGIS v. 3.12, CSR ETRS89 (QGIS Association, http://www.qgis.org); background topographic layer (both main and miniature) from the Norwegian Mapping Authority (https://www.kartverket.no) CC BY 4.0](illustrations/NBR_GrazersMap.png){#fig-sitemap}

## Data collection

We visited each of the 29 sites once between mid-June and late July, progressing from west to east according to respectively earlier and later spring bloom. On each site, we defined a nested sampling design within a representing 400 m^2^ area, preferentially as a 20 x 20 m^2^ square but when not possible (one site) as a 40 x 10 m^2^ area instead, avoiding rocks, waterlogged ground, trees and field edges. The same exclusion rules were applied when setting three plots of 3 x 3 m^2^ within the sampling area, at least 6 m apart from each other and 2 m away from the edge (@fig-sampling). Each plot was divided into nine quadrats of 1 m^2^, five of them dedicated to "non-destructive" plant community collection and the remaining four used for "destructive" soil and beetle sampling.

![Nested sampling design set on a representative area of each grazing field. The 400 m^2^ area was delimited by fences to prevent animals from entering. All data collections were done the same day except from beetles. Credits: figure created on Inkscape v. 1.1 (Inkscape Developers, https://inkscape.org)](illustrations/NBR_GrazersSamplingDesign.png){#fig-sampling}

### Plant and beetle assemblages

All vascular plants were identified to the species level and attributed an average percent cover over each 1 m^2^ quadrat (15 quadrats per site). Identifications were made by the same two observers both years, using the Nordic flora [@mossbergGyldendalsStoreNordiske2018]. Species which could not be identified in the field were coded, collected, and identified *a posteriori* with the help of experts. In the same quadrat, we estimated the percent cover of bare ground, rock, litter, bryophytes and vascular plants. We also estimated sward height, defined as the average height of plant dominant assemblage, at three random locations within the quadrat.

We collected beetles with dung-baited pitfall traps set approx. 1 m from each other in each plot (12 traps per site). The traps were made of white plastic cups of 1 L (diam. 10.5 cm, height 11.5 cm) filled to 1/4 with salted water and a drop of neutral detergent to break water surface tension. We covered them with a square metal grid of 2 x 2 cm mesh size, on which we put a ball of about fist size of previously frozen cattle dung. The dung ball was placed at the centre of the grid, at least 3 cm from the edge of the cups to provide enough space for the beetles to fall in. The traps were left on site for one full week before retrieval. Beetles were sorted at the family level using @duffBeetlesBritainIreland2012a and stored in 70% ethanol. Spatial clumping of the pitfall traps, which compensated for short sampling time, should not have affected the beetle catches [@bakerEvaluatingSpatialAutocorrelation2006; @wardEffectsPitfallTrap2001].

### Soil data

We characterised soil structure with three complementary measurements: bulk density (BD) as a proxy for soil compaction, soil penetration resistance induced by root mat density and soil texture, and gravimetric water content (GWC) for water holding capacity. We measured soil penetration resistance by dropping a sharpened metal rod of 43.4 cm length (diam. 2.2 cm; weight 1.3 kg) from 1 m above ground through a PVC tube to ensure vertical fall. The visible standing part of the stick was measured and subtracted from the total stick length to obtain the length of penetration into the soil. Before dropping the stick, the surrounding vegetation around a 5 cm radius was cleared to prevent resistance from the aboveground vegetation. We performed two penetration tests per "destructive" quadrat, for a total of 24 tests per site. The tip was regularly sharpened and remeasured, and we observed a 1.6 % decrease in length (0.7 cm) by the end of the sampling season.

For bulk density and gravimetric water content measurements, we sampled three soil cores with PVC tubes of 57.23 cm^3^ (diam. 3.7 cm, height 10 cm) in each "destructive" quadrat (36 cores per site). To limit soil compression when hammering down the tubes, we sharpened the edges beforehand and used a knife to cut the surrounding root mat. Samples were covered with plastic foil and stored at 6°C before processing in the lab. If necessary, the soil volume of each core was corrected for holes or unfilled space prior to weighing. As actual soil volume could be low due to sampling constraints (high root density, shallow or sandy soil, steep slope), we discarded cores with less than 50 cm^3^ of soil to prevent inaccurate measurements of bulk density and soil moisture, which represented approx. 20% of the total collected cores.

Gravimetric water content is calculated as the difference in proportion between the soil fresh weight and the soil dry weight, and bulk density as the soil dry weight related to its volume:

$$
\text{Gravimetric water content } (\text{%}) = \frac{\text{Core moist weight } (g) - \text{Core dry weight } (g)}{\text{Core moist weight }(g)} *100
$$

$$
\text{Bulk density }(g/cm^3) = \frac{\text{Core dry weight }(g)}{\text{Core volume }(cm^3)}
$$ 

Usually, moist weight is measured from freshly sampled soil. However, as weather conditions varied considerably during our sampling, we standardised the measurement by weighing soil 48h after water saturation. Dry weight was obtained after 24h of drying in the oven at 105°C. After data filtering of low volumes and some negative values linked to a weighing balance dysfunction during lab processing, all sites had at least 9 measurements of bulk density and gravimetric water content.

For chemical analysis, we sampled soil with a stainless steel cylindrical core of 200 cm^3^ (diam. 10.5 cm; height 12 cm) in each "destructive" quadrat (12 cores per site). The soil was stored at 4°C before being sieved and pooled at the plot level. We sent 0.5 L of each pooled sample to a certified analysis laboratory to determine total nitrogen content with modified Kjeldahl method (NS-EN 13654-1:2890) [@bremnerDeterminationNitrogenSoil1960], and available phosphorus concentration with ammonium lactacte extraction method [@brayDeterminationTotalOrganic1945].

### Environmental data

We recorded field slope angle, elevation and field coordinates on site with a compass clinometer and a GPS (SCR EPSG:25832). We determined aspect degree from a Digital Elevation Model of 10 m x 10 m cell size retrieved from the GeoNorge online map catalog of the Norwegian Mapping Authority [@kartverketGeoNorgeKartkatalogen2023]. From the same resources, we also used a WMS Digital terrain model to measure distance to closest sea or fjord bodies. Average annual precipitation, July maximum temperatures and January minimum temperatures were calculated from 1 km^2^ daily projections provided by the Norwegian Centre for Climate Services over the period 2010-2020 [@wongGriddedKmClimate2016]. For sites not covered by the raster layer, values were determined from averaging values of neighbouring cells with the R software package *terra* (v. 1.7.29) [@hijmansTerraSpatialData2023; @rcoreteamLanguageEnvironmentStatistical2023].

For surrounding landscape composition, we calculated land use percent cover in a 1 km radius area around each field from land cover data provided by the Norwegian Institute of Bioeconomy Research (NIBIO) [@norskinstituttforbiookonominibioKildenArealinformasjon2023]. Land cover was categorised according to the "areal type" AR5 classification system [@ahlstromAR5Klassifikasjonssystem2019], that we simplified into the following categories: forest area, including managed and non-managed; cultivated area, either fully or partially; infield area, i.e. productive grazing area (permanent grasslands); outfield area, i.e. non-productive grazing area (shrublands or heathlands); wetland area; artificial area.

### Field management data

We conducted phone interviews with farmers from October 2021 to January 2022 to collect information on field management. For the 3 out of 26 farmers who were unable or unwilling to reply to our survey, we used data from a similar interview series conducted in 2018 [@reiersenBaerekraftVestnorskLandbruk2019]. Regarding grazing management, we asked the farmers for the following elements: the type of livestock which has been grazing the site for the past 5 years; the size of the flock, including adults and young; the number of months the animals stayed on the field; the total surface area of infields grazed by the animals on the farm; the size of the grazing field we selected as a survey site.

Standard stocking density is calculated as following: 

$$
\text{Stocking density} = \frac{\text{number animals}\times{\text{standard livestock unit}}}{\text{area}\times{\text{grazing period}}}
$$ 

However, estimating stocking density for rotational grazing systems presented some challenges. Usually, farmers in Western Norway move their livestock across different fields year round, according to the productivity or state of the fields at any given point in time, without following a consistent schedule. Thus, as most respondents could not give us accurate grazing period for the infield we selected, we arranged the formula to calculate an average stocking density for the farms' total infield area over a year:

$$
\text{Average stocking density} = \frac{\text{Nadults}\times{\text{standard  LU}} + \text{Nyoung}\times{\text{young LU}} }{\text{total infield area of the farm}}
$$ 

As to our knowledge there is no standard livestock unit for young sheep and goats, we estimated it as half the value of adult (0.05 LU). Total infield area was retrieved from the map of agricultural properties made by the Norwegian Institute of Bioeconomy Research (https://gardskart.nibio.no/). We gathered in this category areas classified as infields (*innmark*), but also unclassified areas which were known as grazing fields from the farmer and/or from field observation. For two farms (IG1 and IG2) which livestock was moved to mountain summer farms every year, we calculated average stocking density for the time period when animals were in the lowlands.

## Data analysis

Canonical correlation analysis searches for the maximum correlation between linear projections of two datasets [@hotellingRelationsTwoSets1992]. We performed four canonical correlation series to investigate 1. the cascading influence between explanatory sets of environmental and management variables, and 2. the influence of these different sets on dominant grass, forb and beetle assemblages. All statistical analyses were performed on R (v.4.4.1) with R-Studio (v.2023.12.1+402) [@positteamRStudioIntegratedDevelopment2023; @rcoreteamLanguageEnvironmentStatistical2024]. The canonical correlation analyses and the related statistical tests were made with the *CCA* [@gonzalezCCACanonicalCorrelation2023] and *CCP* [@menzelCCPSignificanceTests2022] packages. The redundancy analyses were performed with the *vegan* package [@oksanenVeganCommunityEcology2024] and their visual representations were made with the *ggord* package [@beckGgordOrdinationPlots2022]. All data from nested sampling design were averaged at the site level.

Explanatory variables were transformed accordingly to address the method requirements and the high heterogeneity of units: livestock type was coded into dummy numeric and all variables were centered and scaled. We verified and filtered our explanatory variables first according to raw data distribution, removing variables with highly skewed distributions or with obvious outliers, and second by testing variable colinearity within each set with paired plots. For a correlation coefficient of 0.8 or higher, variables were selected according to within site replication and/or overall distribution if applicable. The variables kept for the canonical correlation analysis are shown in @fig-fjordconcept and their statistical summary can be found in Supplementary Information (@tbl-varsum).

Assemblage data were percent cover for grass (Poaceae) and forb species, and abundance for beetle families. We classified taxa according to their frequency and abundance across sites: species or families found in at least one-third of the infields, with a minimum average percent cover of 1% for the former and a minimum average occurrences of 3 for the latter, were selected for the analysis. In total, eight species of grasses, seven species of forbs and five families of beetles were selected (@fig-fjordconcept). Community data were transformed using Hellinger transformation, which lowers the weight of variables with low counts and many zeros [@borcardNumericalEcology2011; @legendreEcologicallyMeaningfulTransformations2001].

We tested for statistical significance of correlations with asymptotic approximation of Hotelling-Lawley Trace (*p.asym* function in CCP package) [@mullerPracticalMethodsComputing1984]. Significant canonical correlations were investigated further with Redundancy Analyses (RDA) [@legendreNumericalEcology2012]. We used adjusted R^2^ [@legendreTestingSignificanceCanonical2011; @peres-netoVariationPartitioningSpecies2006] and tested statistical significance of RDA models, axes and terms by ANOVA-like permutation (n = 999) [@legendreTestingSignificanceCanonical2011; @legendreNumericalEcology2012]. For illustrative purpose, we performed RDAs on all combinations of sets and provided them in Supplementary Information (@tbl-rdaenv; @tbl-rdagrass; @tbl-rdaforb; @tbl-rdabeetle).

<!-- 3) Linearity assumption between explanatory sets and community data; we performed Tukey tests on linear model residuals between community ordination scores and individual explanatory variable [**REF**]. Ordination scores were extracted either from preliminary Detrended Correspondance Analysis (DCA), or from Principal Component Analysis (PCA) if the DCA's first axis length was under 2 units.  -->
<!-- - RDA: multivariate regression of response matrix on explanatory matrix, coupled with a principal component analysis on resulting vectors of fitted values  -->

# RESULTS

We collected a total of 45685 beetles belonging to 14 families, from which 25161 Staphylinidae, 11336 Ptiliidae, 6322 Hydrophilidae, 1128 Scarabaeidae, 1058 Carabidae.

![Summary of canonical correlation series on cascading influence between the four explanatory variable sets (A) and between explanatory sets and dominant grass (B), forb (C) and beetle (D) assemblages. Correlation coefficient rho is provided for the first dimension per default and for all other significant dimensions. Transparent symbols represent non-significant relationships, and arrow thickness is proportional to the rho value. Significance levels are: ° for p-value \< 0.1; \* for p-value \< 0.05; \*\* for p-value \< 0.01; \*\*\* for p-value \< 0.001.](illustrations/NBR_CancorResults.png){#fig-cancor}

::: {#tbl-cancorpanel layout-ncol="2"}
```{r CancorEnv}
#| label: tbl-cancorenv
#| tbl-cap: "Between explanatory sets."
#| message: false
#| warning: false

library(gt)
library(tidyverse)

cancorenv <- subset(envar, select = -c(id, stat))
cancorenv |> 
  gt(groupname_col = "explanatory", rowname_col = "response") |> 
  cols_move_to_start(
    columns = c("rho", "dim")
  ) |> 
  cols_move(
    columns = rho,
    after = dim
  ) |> 
  cols_label(
    #stat = "statistic",
    approx = "F-approx",
    p.value = "p-value",
    dim = "dimension"
  ) |> 
  fmt_number(
    columns = c("approx", "rho"),
    decimals = 2
  ) |> 
  fmt_number(
    columns = p.value,
    decimals = 4
  ) |> 
  opt_stylize(style = 6, color = 'gray')

```

```{r CancorCom}
#| label: tbl-cancorcom
#| tbl-cap: "Between explanatory sets and dominant assemblages of grass, forb and beetle."
#| message: false
#| warning: false

library(gt)
library(tidyverse)

cancorcom <- subset(comvar, select = -c(id, stat))
cancorcom |> 
  gt(groupname_col = "explanatory", rowname_col = "response") |> 
  cols_move_to_start(
    columns = c("rho", "dim")
  ) |> 
  cols_move(
    columns = rho,
    after = dim
  ) |> 
  cols_label(
    #stat = "statistic",
    approx = "F-approx",
    p.value = "p-value",
    dim = "dimension"
  ) |> 
  fmt_number(
    columns = c("approx", "rho"),
    decimals = 2
  ) |> 
  fmt_number(
    columns = p.value,
    decimals = 4
  ) |> 
  opt_stylize(style = 6, color = 'gray')

```

Summary of statistical tests for canonical correlations with F approximation of Hotelling-Lawley Trace.
:::

![Redundancy analysis (RDA) plots for significant canonical correlations between fjord and landscape (A), grazing and fine-scale environment (B), fine-scale environment and dominant grass assemblage (C), and fine-scale environment and dominant forb assemblage (D). Abbreviations: AnnPreci: annual precipitation; Elev: elevation; JulTemp: average July temperature; SeaDist: distance to sea; StockDens: average stocking density; Bryo: percent bryophyte cover; BD: bulk density; SoilPene: soil penetration](outputs/RDAresults.png){#fig-rda}

::: {#tbl-rdapanel layout-ncol="2"}
```{r RDAexp}
#| label: tbl-rdaexp
#| tbl-cap: "Between explanatory sets."
#| message: false
#| warning: false

library(gt)
library(tidyverse)

rdaexp <- filter(rdastat_all, model == "FjordxLandscape" | model == "GrazingxFinescale")
rdaexp <- subset(rdaexp, select = -c(r.squared, type, eigenval, response, explanatory))
rdaexp |> 
 # gt(groupname_col = "explanatory", rowname_col = "response") |> 
 # tab_options(row_group.as_column = TRUE) |> 
 gt(groupname_col = "model", rowname_col = "dim") |> 
  cols_move_to_start(
    columns = c("dim", "adj.r.squared")
  ) |>
  cols_move(
    columns = Df,
    after = F
  ) |>
  cols_label(
    Variance = "variance",
    PrF = "p-value",
    Df = "df",
    adj.r.squared = "adjRsq"
  ) |>
  fmt_number(
    columns = c("Variance", "F", "adj.r.squared"),
    decimals = 2
  ) |>
  fmt_number(
    columns = PrF,
    decimals = 4
  ) |>
  sub_missing(missing_text = "-") |> 
  opt_stylize(style = 6, color = 'gray')

```

```{r RDAcom}
#| label: tbl-rdacom
#| tbl-cap: "Between explanatory sets and dominant assemblages of grass, forb and beetle."
#| message: false
#| warning: false

library(gt)
library(tidyverse)

rdacom <- filter(rdastat_all, model == "FinescalexGrass" | model == "FinescalexForb")
rdacom <- subset(rdacom, select = -c(r.squared, type, eigenval, response, explanatory))
rdacom |> 
 # gt(groupname_col = "explanatory", rowname_col = "response") |> 
 # tab_options(row_group.as_column = TRUE) |> 
 gt(groupname_col = "model", rowname_col = "dim") |> 
  cols_move_to_start(
    columns = c("dim", "adj.r.squared")
  ) |>
  cols_move(
    columns = Df,
    after = F
  ) |>
  cols_label(
    Variance = "variance",
    PrF = "p-value",
    Df = "df",
    adj.r.squared = "adjRsq"
  ) |>
  fmt_number(
    columns = c("Variance", "F", "adj.r.squared"),
    decimals = 2
  ) |>
  fmt_number(
    columns = PrF,
    decimals = 4
  ) |>
  sub_missing(missing_text = "-") |> 
  opt_stylize(style = 6, color = 'gray')

```

Summary of RDA permutation tests for relationships which showed significant canonical correlations
:::

``` {r InlineNumberFormat}

# Associate values to format class for inline citation
class(cancorenv$rho) <- "stat"
class(cancorenv$p.value) <- "p.value"

# does not work yet...

```

## Cascading influence between explanatory sets

There was a strong canonical correlation between fjord features and landscape composition (first dimension correlation coefficient rho1 = `r envar$rho[1]`, F approximation of Hotelling-Lawley trace F~24,70~ = `r envar$approx[1]`, p-value = `r envar$p.value[1]`) (@fig-cancor A; @tbl-cancorenv), which was supported by the RDA model (proportion explained variance = `r rdastat_all$Variance[1]`, F~4~ = `r rdastat_all$F[1]`, p-value = `r rdastat_all$PrF[1]`) (@tbl-rdaexp). The most important factors was annual precipitation (var = `r rdastat_all$Variance[4]`, F~1~ = `r rdastat_all$F[4]`, p-value = `r rdastat_all$PrF[4]`), with more cultivated lands and infields in wetter areas (@fig-rda A). There was no other significant canonical correlation between explanatory sets (@fig-cancor A; @tbl-cancorenv), but the RDA model between grazing and the fine-scale environment was nevertheless marginally significant (var = `r rdastat_all$Variance[6]`, F~5~ = `r rdastat_all$F[6]`, p-value = `r rdastat_all$PrF[6]`) with a stronger effect of livestock (var = `r rdastat_all$Variance[9]`, F~1~ = `r rdastat_all$F[9]`, p-value = `r rdastat_all$PrF[9]`) and field size (var = `r rdastat_all$Variance[10]`, F~1~ = `r rdastat_all$F[10]`, p-value = `r rdastat_all$PrF[10]`) (@tbl-rdaexp). As such, bigger fields and/or grazed by sheep tend to have higher cover of bryophytes and more compact soil but lower vegetation height (@fig-rda B).

## Cascading influence on dominant grass, forb and beetle assemblages

There were quite a few interactions between environmental gradients and grass and beetle dominant assemblages which showed at least marginally significant in canonical correlation analyses (@fig-cancor; @tbl-cancorcom) but not in the corresponding RDA (SI: @tbl-rdagrass; @tbl-rdabeetle), even if the landscape x beetle model still explained a fair share of variance. Only the relationships between plant dominant assemblages and the fine-scale environment were found significant for both the canonical correlations and the redundancy analyses (@fig-cancor; @tbl-rdacom). The fine-scale environment x grass RDA model explained 48% of the variance (R~adj~^2^ = `r rdacom$adj.r.squared[1]`, F~8~ = `r rdacom$F[1]`, p-value = `r rdacom$PrF[1]`), with the two first axes accounting for 26% (F~1~ = `r rdacom$F[3]`, p-value = `r rdacom$PrF[3]`) and 10% (F~1~ = `r rdacom$F[4]`, p-value = `r rdacom$PrF[4]`) respectively (@tbl-rdacom). The first axis showed a vegetation gradient opposing taller sward (var = `r rdacom$Variance[6]`, F~1~ = `r rdacom$F[6]`, p-value = `r rdacom$PrF[6]`), with associated species *Deschampsia cespitosa* and *Holcus lanatus*, to well-developed bryophyte layer (var = `r rdacom$Variance[5]`, F~1~ = `r rdacom$F[5]`, p-value = `r rdacom$PrF[5]`), with associated species *Poa pratensis* (@fig-rda C). The second axis was driven by an edaphic gradient, with soil acidity (var = `r rdacom$Variance[7]`, F~1~ = `r rdacom$F[7]`, p-value = `r rdacom$PrF[7]`) associated with low resistance to penetration (var = `r rdacom$Variance[8]`, F~1~ = `r rdacom$F[8]`, p-value = `r rdacom$PrF[8]`) and favouring *Deschampsia flexuosa* over *Festuca rubra* and *Holcus lanatus*. The RDA model between fine-scale environment and dominant forb assemblage explained 39% of the variance (R~adj~^2^ = `r rdacom$adj.r.squared[9]`, F~8~ = `r rdacom$F[9]`, p-value = `r rdacom$PrF[9]`) and showed fairly similar gradients (@fig-rda D) but with lower explanatory power (@tbl-rdacom). The first axis accounted for 18% of the variance (F~1~ = `r rdacom$F[11]`, p-value = `r rdacom$PrF[11]`) and was mainly driven by soil penetration (var = `r rdacom$Variance[13]`, F~1~ = `r rdacom$F[13]`, p-value = `r rdacom$PrF[13]`), inversely related to soil pH and available phosphorus concentration (SI: @tbl-rdaforb). Seemingly less-productive soils with lower resistance were associated with *Potentilla erecta* and *Galium saxatile*, while more productive and compact soils were associated with *Ranunculus acris* and *Trifolium repens*.

<!-- Canonical correlation analyses showed at least marginally significant relationships between grass dominant assemblage and fjord-scale variation (`r comvar$rho[1]`, F~28,66~ = `r comvar$approx[1]`, p-value = `r comvar$p.value[1]`) as well as grazing (rho1 = `r comvar$rho[3]`, F~35,77~ = `r comvar$approx[3]`, p-value = `r comvar$p.value[3]`), and between beetle dominant assemblage and fjord-scale variation (`r comvar$rho[11]`, F~20,74~ = `r comvar$approx[11]`, p-value = `r comvar$p.value[11]`), landscape (rho1 = `r comvar$rho[10]`, F~30,82~ = `r comvar$approx[10]`, p-value = `r comvar$p.value[10]`) and grazing (rho1 = `r comvar$rho[13]`, F~25,87~ = `r comvar$approx[13]`, p-value = `r comvar$p.value[13]`), but none were supported by corresponding RDA even though the landscape x beetle model still explained a fair share of variance (@tbl-rdabeetle) (SI: @tbl-rdagrass; @tbl-rdabeetle). -->


# DISCUSSION

The interaction between environmental gradients across scales was rather weak as we only observed a single well-supported correlation between the regional-scale gradient and the landscape composition, with an increase of the proportion of crops and infield towards wetter inland area. Regarding assemblage responses, there was a significant influence of fine-scale environment on dominant grass and forb species, mainly driven by sward height, soil resistance to penetration and bryophyte cover. However, we found no clear evidence that plant and beetle assemblages were affected by other environmental gradients as canonical correlations and redundancy analyses showed contradictory results.

## Site productivity and edaphic factors affect plant dominant assemblages

Common grass and forb species were rather affected by fine-scale environmental variation than variables associated with grazing management, landscape and fjord-driven regional gradient [@hejcmanRengenGrasslandExperiment2010; @myklestadSoilSiteManagement2004; @siefertScaleDependenceVegetationenvironment2012]. Both assemblages responded to two similar gradients: 1) sward height against bryophyte cover and 2) soil resistance to penetration against soil acidity, explaining the highest variation for grasses and forbs respectively.

Sward height, which depends on individual species' growth form, can only be a poorly reliable predictor of species distribution when considered on its own, unless in strictly controlled experimental conditions [e.g. see @marriottLongtermImpactsExtensive2009 and @tonnGrazingInducedPatchiness2019]. For instance in our study the tall and unpalatable *Deschampsia cespitosa*, where dominant, most probably contributed to increase average vegetation height. However, when inversely related to bryophyte cover and, still consistently even if not significantly, to slope, it may be indicative of an underlying productivity gradient [@grimeCompetitiveExclusionHerbaceous1973; @hejcmanRengenGrasslandExperiment2007; @olffEffectsHerbivoresGrassland1998; @vanderwalMossesStruggleLight2005] with well-drained fields at one end and flatter, more productive fields at the other [@losvikHayMeadowCommunities1993]. Slope angle can indirectly affect both plant height [@dorroughPlantNeighbourhoodLandscape2007] and species composition [@klimekPlantSpeciesRichness2007] through for instance the depletion of soil nutrients [@kleinebeckerEffectsGrazingSeasonal2011] or the change of animal grazing behaviour [@brownImportanceScaleEvaluating1989; @fynnManagingBrowsingGrazing2019; @hesterForagingBehaviourSheep1999], but its effect is likely to be mediated by the slope aspect [@bennieInfluenceSlopeAspect2006; @dauberLandscapeStructureIndicator2003;  @turtureanuScaleTaxondependentBiodiversity2014]. The combined influence of slope angle and aspect can be assessed through the heat load index (HLI), which approximates the amount of solar energy directly hitting the ground and thus available for plant growth [@merunkovaEnvironmentalControlSpecies2012; @turtureanuScaleTaxondependentBiodiversity2014]. Though this measure was initially considered in our analysis it was discarded because of obvious outliers, and instead we included slope angle and aspect as independent variables.

Soil resistance to penetration depends not only on soil properties including structure, texture, water content and carbon ratio, but also on biotic components such as plant roots [@forsterRootTraitsCrop2022; @mirrehInfluenceSoilBulk1972; @sojkaSituStrengthBulk2001]. Correlation with one or the other parameter is however far from being straightforward, as for example, soil resistance has been found to be either strongly [@gabrielCoverCropsReduce2021; @negronConsequencesIntensiveGrazing2019], weakly [@kumiQuantifyingRootinducedSoil2023] or conditionally [@sojkaSituStrengthBulk2001] associated with water content, depending on the considered soil layer. Likewise, average plant root diameter and seemingly root volume tend to increase soil resistance in experimental conditions, but the consistency of observations would be highly variable depending on soil texture and plant species [@kumiQuantifyingRootinducedSoil2023]. In our study, soil resistance to penetration was independent from bulk density, and by extent gravimetric water content and loss of ignition, but was systematically paired with a pH gradient. Lower resistance on more acidic soils could be explained by the increased concentration of ammonium [@brittoNH4ToxicityHigher2002; @kleijnSearchKeyBiogeochemical2008] and ionic aluminium [@halingRootMorphologyRoothair2010; @kochianCellularMechanismsAluminum1995] which would inhibit plant root growth. However, experimental acidification in grasslands showed the opposite trend with an increase of root density following a community shift from forbs to perennial grasses [@chenBioticCommunityShifts2015; @wangSoilAcidificationAlters2020]. We cannot extrapolate further on this soil acidity-resistance relationship without direct assessment of root length/density and standardised measurement of soil resistance to penetration [@sojkaSituStrengthBulk2001], but this interaction would definitely be worth investigating in further research.

In contrast with results from both long-term experiments [@avolioChangesPlantCommunity2014; @hejcmanRengenGrasslandExperiment2010; @kiddImpacts120Years2017; @koernerNutrientAdditionsCause2016; @zarzyckiSchemeNutrientAddition2020] and observational studies [@critchleyAssociationLowlandGrassland2002; @klimekPlantSpeciesRichness2007; @raatikainenImpactsLocalRegional2007; @rupprechtEffectsYearroundGrazing2016] we found no significant influence of soil available phosphorus and total nitrogen on dominant grass and forb assemblages. Plant growth is not only affected by macro-nutrients [@fayGrasslandProductivityLimited2015], their chemical forms [@ceulemansPhosphorusResourcePartitioning2017] or application modes (e.g. amount, frequency and timing) [@rosOptimalUsePhosphorus2020; @sittersImpactsBrowsingGrazing2019], but also by their interactions with the existing soil nutrient pool [@bindrabanRevisitingFertilisersFertilisation2015], pH [@caiIntensifiedSoilAcidification2015; @kiddImpacts120Years2017; @mengGlobalSoilAcidification2019] and microbial community [@dincaFertilizationSoilMicrobial2022; @farrerTeasingApartPlant2016; @leechComparativeEffectAlternative2019], making individual species response to nutrient addition challenging to detect [@lapierreDriversVariationAboveground2016]. Total nitrogen, though widely used as indicator of fertilisation regimes for practical and cost-efficiency reasons [@hejcmanRengenGrasslandExperiment2010; @myklestadSoilSiteManagement2004], offers poor approximation of the actual concentrations of inorganic, available forms for plants i.e. nitrate or ammonium [@collinsSoilNitrateConcentrations2004; @stluceNitrogenDynamicsIndices2011]. On the other hand Bray's phosphorus extraction method should show relatively accurate estimation of phosphorus uptake by plants [@gilbertAvailableSoilPhosphorus2009]. However, relying on this gradient alone undoubtedly limited the study outputs [@kiddImpacts120Years2017; @zarzyckiSchemeNutrientAddition2020], especially considering the wide variety of fertilisation practices reported by the interviewed farmers.

<!-- - variance of floristic composition second best explained by pH after elevation [@zelnikPlantSpeciesDiversity2013] -->
<!-- In managed grasslands, soil acidity mainly depends on bedrock type [@balstromStatisticallyBasedMapping2013; @fabianGEMASSpatialDistribution2014] and type, amount and/or frequency of fertilisation [@leechComparativeEffectAlternative2019; @vasakSoilPHChanges2015]. Bedrock type was not included in our analysis due to poor replication of the different categories in our set of sites, but *a posteriori* we did not observe clear trends in pH between soils from more acidic granitic gneiss (pH ranging from 4.3 to 5.0) and from more basic anorthosite (pH ranging from 4.9 to 5.1) [@norgesgeologiskeundersokelseNasjonalBerggrunnsdatabase2023]. -->
<!-- - Regarding soil pH for instance, mineral fertilisers such as inorganic nitrogen can dramatically acidify the soil [@kiddImpacts120Years2017], while animal manure and lime are usually used as neutralisers [@caiIntensifiedSoilAcidification2015; @hejcmanRengenGrasslandExperiment2010; @kalkhoranOptimalLimeRates2020; @xuModelbasedOptimalManagement2023]. -->
<!-- - + narrow time window which did not take in account farmer's time of fertilisation + seasonal variation of nutrient patchiness [@ryelTemporalDynamicsSoil1996] -->


## Looking for the dung beetles in the dung

The distribution of beetle main families was never correlated with the fine-scale environment of the NBR grazing fields, even though vegetation structure [@broseBottomupControlCarabid2003; @harveyEffectsChangesPlant2008; @ngDisentanglingEffectsFarmland2018; @pratherPlantsRegulateGrassland2019] and soil moisture [@buchholzSimulatingSmallScale2013; @lafageRelativeImportanceManagement2016] can both shape arthropod communities by affecting the ecological niches [@pratherPlantsRegulateGrassland2019], trophic interactions [@wiseImpactsRainfallExtremes2019] and life cycles [@hollandSpatiotemporalDistributionEmergence2007; @torodeAlteredPrecipitationImpacts2016]. Family level may have been too coarse to detect small-scale changes within the relatively tight habitat range that are Nordic semi-natural wet meadows, as 55% of our catches were made of Staphylinidae which feature very high functional, morphological and ecological diversity and encompass many stenotopic species [@bohacStaphylinidBeetlesBioindicators1999; @irmlerEffectEnvironmentalConditions2018]. Identification at the species level however requires both substantial amount of time and high level of expertise which are challenging to meet in short-term multitaxa studies. More accessible alternatives are trait-based approaches [@neffCrossscaleEffectsLand2019; @weissUnifyingFunctionalTrait2019] that also provide both fundamental knowledge on ecological interactions and applied prospects for habitat monitoring [@nashDiscontinuitiesCrossScale2014; @morettiHandbookProtocolsStandardized2017; @wongTraitBasedEcology2019]. A suitable indicator would be for instance body size [@burelDifferentialResponseSelected2004; @ngDisentanglingEffectsFarmland2018; @tonelliEffectsProgressiveAbandonment2018], which assessment relies on well-established and easily replicable methodology, and which relationships with numerous biological and ecological features such as metabolism, trophic level, feeding strategies, stress-tolerance or ecological niches have been thoroughly investigated [@labarberaAnalyzingBodySize1989; @osorio-canadasBodySizePhenology2016; @petersEcologicalImplicationsBody1993; @woodwardBodySizeEcological2005].

Interestingly, our samples were largely dominated by the Staphylinidae, Hydrophilidae and Ptiliidae families that do not belong to the "dung beetle" group but nevertheless encompass dung-associated species [@daoudiCompositionPhenologyBeetle2022; @hanskiDungInsectCommunity1991], while the Scarabaeidae family only represented 2.5% of our total catches. Widely used as indicators of grassland functionality, the paraphyletic dung beetle group is primarily characterised by the species coprophagy and their subsequent role in the nutrient cycling process between the livestock, the soil system and the vegetation [@nicholsEcologicalFunctionsEcosystem2008]. However the absence of rigorous taxonomic definition raises some issues: first, we observe a generalised lack of consistency regarding the dung beetle terminology, which can indiscriminately includes the single Scarabaeinae (Scarabaeidae) [@barraganImpactGrazingDung2014; @nicholsEcologicalFunctionsEcosystem2008; @simmonsEcologyEvolutionDung2011], the Aphodiinae (Scarabaeidae) and the Geotrupidae [@buseRelativeImportancePasture2015; @finnReviewCompetitionNorth2003; @tonelliEffectsProgressiveAbandonment2018], or even the whole Scarabaeidae family [@arellanoDungBeetlesColeoptera2023]; second and more importantly, *a priori* reducing the coprophagous community to the above-listed taxa when the trait is shared with other families (e.g. Hydrophilidae or Staphylinidae) [@hanskiDungInsectCommunity1991] could create critical research biases [@tonelliEffectsProgressiveAbandonment2018] and knowledge gaps, particularly in the Nordic region where Scarabaeinae & Aphodiinae are rather poorly represented [@ArtsdatabankenKunnskapsbankNaturmangfold2023]. Investigating the overall dung-associated community would undoubtedly require considerably more time and expertise, but it should nevertheless be encouraged [@daoudiCompositionPhenologyBeetle2022] especially with the tremendous potential offered by DNA-based methods [@sigsgaardEnvironmentalDNAMetabarcoding2021].

<!-- - @lichtenbergGlobalSynthesisEffects2017 -->
<!-- - @toccoPastoralPracticesReverse2013 -->
<!-- - carabid community in field relatively stable after 10 years [@purtaufCarabidCommunitiesSpatiotemporal2004] -->
<!-- - distribution carabid community along forest-grassland landscape gradient [@vanbergenEffectLanduseHeterogeneity2005] -->
<!-- - ground beetle ecological preferences affected by the dominant grazing system in the landscape matrix [@martelImpactFarmingSystems2019] -->
<!-- - taxa (butterflies, carabids, plants) respond differently to landscape features and landscape more important driver for high mobility organisms [@weibullSpeciesCompositionAgroecosystems2003] -->
<!-- - generalist beetle richness and abundance negatively correlated to the % cover grassland within 500 m radius around field [@bataryResponsesGrasslandSpecialist2007] -->

## Isolated influence of landscape and fjord-driven gradients

We observed that the current agro-ecological landscape in the NBR still partially reflects the land use heritage from the past century with extensive grazing on the flat, low-productive and climatically buffered coastline and more productivity-driven management inland [@hamreLandcoverStructuralChanges2007a; @kalandNordhordlandBiosphereReserve2018]. However, neither the surrounding habitat composition or the location along the fjord system showed clear effects on the farming, abiotic conditions or dominant assemblages of the grazing fields. In the absence of strong elevation gradient [@kaltsasGroundBeetlesRespond2013; @buzhdyganDirectIndirectEffects2020], drastic biogeographical discontinuity [@baldiEffectsGrazingBiogeographic2013; @neffCrossscaleEffectsLand2019] and heavy land use intensification [@burelDifferentialResponseSelected2004; @schweigerQuantifyingImpactEnvironmental2005; @weibullSpeciesCompositionAgroecosystems2003] in our study area, the magnitude of environmental variation may have been too low to affect the most common taxa [@clelandSensitivityGrasslandPlant2013]. The presence of woodlands or wetlands around the field may be a source of non-grassland specialists, but it would not be reflected amongst dominant assemblages [@avironCarabidAssemblagesAgricultural2005]. More importantly, our study relied on an exploratory, highly simplified design that used unsuitable spatial resolution and model structure for capturing broad-scale processes. First, we investigated plant and beetle response at a single and relatively small scale (1 m^2^), which would only weakly reflect the climate variation driven by spatial heterogeneity [@dixonhamilCrossscaleContradictionsEcological2016; @graaeStayGoHow2018] and overall lessen its primacy compared with local or fine-scale variables [@siefertScaleDependenceVegetationenvironment2012]. Second, both the canonical correlation and redundancy analyses assume linear relationships between variable sets while these dynamics are not easily verified for cross-scale interactions, especially in a context of high spatio-temporal variability [@dixonhamilCrossscaleContradictionsEcological2016; @nashDiscontinuitiesCrossScale2014; @petersCrossScaleInteractions2007]. These are significant methodological limitations which could have led to underestimate the influence of broader gradients on local and fine scales.

The unique regional context induced by the abrupt collision between oceanic and mountain regimes within the fjord system [@beylichLandscapesLandformsNorway2021; @nesjeFjordsNorwayComplex2009] offers valuable potential to investigate and disentangle the cross-scale dynamics affecting terrestrial habitats and communities [@raatikainenImpactsLocalRegional2007; @reitaluPlantSpeciesSegregation2008; @scherrerDisentanglingProcessesDriving2019]. Such venture would however require much finer characterisation of the fjord-driven environmental gradient along with a thorough examination of the intermediate processes bridging the considerable spatio-temporal gap with the local and fine scales [@petersCrossScaleInteractions2007]. The role of spatial heterogeneity is expected to be rather crucial in shaping cross-scale interactions, from the large seasonal lag constraining the phenology of a whole North-facing fjord's arm side to the micro-climatic niches created by topographic complexity [@graaeStayGoHow2018]. This type of research is challenging and resource-demanding, but it is becoming increasingly relevant in the current context of global changes [**ref**]. In this regard, agro-ecological systems in the lowlands have so far received considerably less attention than alpine and sub-alpine habitats because they are not as vulnerable to shifts in temperature and precipitation regimes [@gyaTestLocalAdaptation2023; @klanderudImportanceBioticVs2015; @vandvikRolePlantFunctional2022]. Still, semi-natural grasslands also display a high ecological diversity and support key ecosystem services for human livelihood [@kremenLandscapesThatWork2018; @bengtssonGrasslandsMoreImportant2019] which are bound to be affected by large-scale changes that could be expected to be even more unpredictable in the fjord system context [@bardgettCombattingGlobalGrassland2021].

<!-- - stronger ecological effects when multiple global change factors/pressures act simultaneously instead of individually [@speisserNumberSimultaneouslyActing2022] -->
<!-- - global regional effect strongest factor of beetle functional group diversity & community composition, but different response according to functional groups [@liuFunctionalBeetleDiversity2014] -->
<!-- - Appropriate indicators/analyses: landscape composition easier indicator but maybe less relevant than configuration [@cousinsHotspotsAreGone2008; @evjuSpatialConfigurationMatters2016; @lecoqLandscapeStructureInfluences2023; @ostmanLandscapeComplexityFarming2001; @schweigerQuantifyingImpactEnvironmental2005], especially in a mosaic-like landscape with low urban & crop elements -->
<!-- - developing research on this kind of model-like mosaic landscape with high connectivity would be beneficial for future research applicable elsewhere in Europe [@grashof-bokdamGreenVeiningLandscape2005] -->
<!-- - Historical dimension: long-term heritage of landscape connectivity on semi-natural grassland species richness [@lindborgHistoricalLandscapeConnectivity2004] -> deep knowledge of the farmers -->

## Assessing grazing impact in small-scale grazing systems

Correlation analyses of grazing management with fine-scale environment and dominant plant and beetle assemblages showed the most inconsistent results, whether be the contradicting outputs between canonical correlation and redundancy analyses or the poor fit of explanatory variables in the ordination space (@fig-rda B). If this makes our study inconclusive regarding the actual effect of grazing management on the NBR grazing fields, it is on the other hand quite representative of how challenging such assessments are in small-scale grazing systems. Under lower disturbance regime and higher management variability the decoupled influences of domestic grazing features (i.e. intensity, timing or grazer type) are more difficult to detect, when they are already far from being consensual, if ever thoroughly investigated, in the context of controlled experiments [@rookMatchingTypeLivestock2004]. For instance grazing pressure, yet considered as the most impactful grazing factor [@dumontPlantHerbivoreInteractions2011], is usually shown to affect grassland communities [@kaltsasGroundBeetlesRespond2013; @kruessContrastingResponsesPlant2002; @ludvikovaLongTermDefoliation2014; @pozsgaiGrazingImpactsGround2022; @skornikEffectGrazingPressure2010] and soil structure [@ludvikovaLongTermDefoliation2014; @zhangSoilPhysicalChemical2019] from a certain threshold only, with no to marginal effects of lower intensity ranges [but see @buzhdyganDirectIndirectEffects2020 and @tonelliEffectsProgressiveAbandonment2018]. Regarding livestock type, the changes in plant species competition and recolonisation dynamics driven by forage uptake and trampling disturbance of animals with different morphology and behaviour are expected to be reflected in the community composition [@grantControlledGrazingStudies1996; @rookMatchingTypeLivestock2004; @wallisdevriesImportanceScalePatchiness1999]. However, if sheep preference for forbs and legumes was shown to affect plant community structure on the long-term [@dumontPlantHerbivoreInteractions2011; @tothLivestockTypeMore2018], most studies reported that shifting to another livestock type [@cuchillohilarioForageSelectivityCattle2018; @fraserImpactSwardComposition2011] or to mixed grazing [@jerrentrupDiverseSwardsMixedgrazing2020; @pakemanLongTermImpacts2019] for up to 15 years did not change the community composition. The sensitivity of plant and arthropod communities to grazing may also not be as straightforward: for example, @baldiEffectsGrazingBiogeographic2013 reported that beetles and not plants responded to grazing intensity, while @ravettoenriBiodiversityfriendlyRotationalGrazing2017 found that beetles were not affected by livestock type and rotational management contrary to butterflies and bumblebees.

Even before considering assemblage responses, there are substantial methodological limitations when attempting to summarise and characterise as accurately as possible the features of the NBR grazing systems, and more particularly the versatility of the rotational management. Rotational grazing is usually investigated in comparison with continuous grazing in experiments with strict control over stocking density and length and timing of grazing period [@boavistaEffectsGrazingRegimes2019; @perottiYearRotationalGrazing2018; @ravettoenriBiodiversityfriendlyRotationalGrazing2017], but such standardisation cannot be met in observational studies of small-scale grazing systems. If in theory each farmer leaves a *certain number of animals*, for a *certain period of time* on a *certain area of land*, in practice there are many "real-life" spatial and temporal dynamics which can derive from either unforeseeable circumstances, such as yearly flock size fluctuation depending on young birth and survival or climatic variation influencing the field productivity, or from the farmer's decision-making, who can, for an arbitrary period of time, split/spread the flock or extend/reduce the grazing field areas. We witnessed it firsthand during our interviews as the farmers themselves struggled to provide accurate grazing schedules for a single field. Such complex rotational dynamics, further emphasized by the lack of farmer ownership over the fields [@forbordDriversChangeNorwegian2014], have also led to distinct management histories: if drastic land use changes were rare amongst our sites (only two with former forestry and crop respectively), we recorded livestock shifts for two-thirds of them, mostly happening during the last 30 years. Considering the lack of consistent information on grazing rotations and the limitations of the canonical correlation analysis, we excluded any form of temporality amongst our grazing management variables but probably at the expense of a fair share of variation resulting from interactions with plant phenology [@lennartssonEffectTimingGrassland2012], arthropod life-cycle [@lenoirEffectsTimingGrazing2010], habitat productivity [@fetzelSeasonalityConstraintsLivestock2017; @gilhausSeasonalVariationsFodder2016] and field history [@marriottLongtermImpactsExtensive2009; @milchunasQuantitativeEffectsGrazing1993].

These shortcomings question whether (comparative) ecological assessments based on single representative fields offer meaningful overviews of the real-life rotational management in small-scale grazing systems. First, standard grazing indicators, even though necessary to ensure research reproducibility, are not tailored to capture the complex livestock flows neither the comprehensive farmer knowledge on both past and current management [@altieriLinkingEcologistsTraditional2004]. Scaling up study designs to encompass the whole farm matrix should provide a more accurate mapping of the livestock flows between grazing fields as well as potential energy transfers with other land uses, that would be necessary for developing novel grazing indicators adapted to small-scale farming context [@deffontainesAgriculturalSystemsLandscape1995; @duelliBiodiversityEvaluationAgricultural1997; @kokEuropeanBiodiversityAssessments2020]. Second, the replication level required for 1) accurately addressing statistical noise brought by high context dependency and 2) relying on suited multivariate analyses to disentangle the various grazing factors from the environment is unattainable for most ecological surveys except wide-scale collaborative projects [e.g. @leprovostSupplyMultipleEcosystem2022]. Instead, aiming for local case studies in which site specificity is inherently part of the design may be even more relevant for small-scale grazing systems, as it would favour direct research application through the experimentation of more innovative, trans-disciplinary approaches and concepts [@wittmanSocialEcologicalPerspective2017] such as farm multifunctionality [@kremenLandscapesThatWork2018; @wilsonWeakStrongMultifunctionality2008] or life cycle assessments [@kokEuropeanBiodiversityAssessments2020]. Farmers are particularly crucial to actively involve as much as possible along the research process [@lacombeDesigningAgroecologicalFarming2018]: they are at the same time knowledge-holders of past and current farm management as well as local conditions [@berkesRediscoveryTraditionalEcological2000; @ianniRevitalizingTraditionalEcological2015], land managers organising the farm structure and management according to socio-economic and environmental constraints [@mcginlayThreatAbandonmentSocioecological2017; @rosa-schleichEcologicaleconomicTradeoffsDiversified2019], and social connectors building community-based knowledge and experience favouring the sustainability and resilience of local farming systems [@lutzFarmerCooperationMeans2017; @shiSocialInteractionEffect2021]. To facilitate the development of knowledge-building capacities amongst small-scale farming communities around the world along with large-scale dissemination of research outcomes, we recommend to rely on already existing global collaboration such as the World Network of UNESCO Biosphere Reserve, which was shown to be an important stepping stone for bridging the gap between science, local communities and policy-making [@barracloughGlobalKnowledgeAction2023].

<!-- -   *Functional resource heterogeneity*, which determines animals accessibility to fodder, is driven both by broader, spatially-fixed or temporally-recurrent factors, such as climate, topography and geology, and by finer, spatially and temporally variable factors, as for instance the patchiness of rainfall, grazing or fertiliser spraying by farmers [@fynnManagingBrowsingGrazing2019]. -->
<!-- - only 15-20% variation in species composition plants, butterflies & carabids explained by all three scale heterogeneity (landscape, farm, field) [@weibullSpeciesCompositionAgroecosystems2003] -->
<!-- - in-field patchiness of vegetation structure with different grazing capacities created by cattle grazing [@gilhausCattleForagingHabits2014] -->
<!-- - limitation using species-based approach only: from savannah example, fodder quality of individual species dependent on rainfall and site productivity, as absence of limiting factors allow plant development towards defensive strategies against herbivores [@hopcraftHerbivoresResourcesRisks2010] -->
<!-- - no influence of rotational cattle grazing scheme on overall plant species composition but differences detected when narrowed at the life trait group in sub-alpine & alpine pastures [@perottiYearRotationalGrazing2018] -->
<!-- - Effects of domestic grazing can also be buffered by initial habitat productivity or diversity [@isselsteinEffectsLivestockBreed2007; @liuImpactsGrazingDifferent2015]. -->
<!-- - cattle vs sheep affected beetle community composition [@lyonsImpactsContrastingConservation2017] -->
<!-- - different plant species composition between cattle rotational and continuous grazing [@boavistaEffectsGrazingRegimes2019] -->
<!-- - A relatively consensual effect of grazing intensity is the reduction of vegetation height [@dorroughPlantNeighbourhoodLandscape2007; @isselsteinEffectsLivestockBreed2007; @pakemanLongTermImpacts2019], even if in our analysis the trend rather non-intuitively showed the opposite -->


# CONCLUSION

In the NBR, the influence of management on the dominant assemblages of semi-natural grasslands seems to be overtaken by the inherent habitat productivity and soil conditions, even if clear relationships are challenging to determine due to the finely entangled interactions between human, environment and livestock that characterise small-scale grazing systems [@bignalLowintensityFarmingSystems1996]. For further research, we emphasized two main directions: 1) the investigation of cross-scale interactions and intermediate processes driven by landform heterogeneity and climate gradient, across different spatial ranges and grains [@petersCrossScaleInteractions2007]; 2) the extensive study of small-scale farming across ecosystems and communities based on tailored farm-level and trans-disciplinary approaches associated with large-scale experience sharing [@barracloughGlobalKnowledgeAction2023]. Considering the considerable ecological burden left by intensive agriculture worldwide, it is an utmost necessity to encourage research on the agro-ecological systems that promote high biological diversity, ensure landscape multifunctionality and local community sovereignty over their food system and land resources [@boavistaEffectsGrazingRegimes2019; @kremenLandscapesThatWork2018; @wangDiversifyingLivestockPromotes2019]

<!-- - We encourage more multi-scale studies investigating the ecological dynamics in the fjord system, investigating different response scales, organism responses [@billeterIndicatorsBiodiversityAgricultural2007; @gonthierBiodiversityConservationAgriculture2014] -->
<!-- - There is a need for more local studies on small-scale grazing or farming systems as the high environmental variation & high diversity of practices make the assessments difficult, especially when involving rotation, for which we encourage socio-ecological approaches [@wittmanSocialEcologicalPerspective2017] -->
<!-- - Socio-ecological approaches bring a breath of fresh air [@bernuesQuantifyingMultifunctionalityFjord2015; @cusensParticipatoryMappingReveals2022; @wittmanSocialEcologicalPerspective2017] -->


# ACKNOWLEDGEMENTS

# STATEMENTS AND DECLARATIONS

## FUNDING

## AUTHOR CONTRIBUTION

-   Conceptualization, A.E.E., M.K.;
-   Methodology, A.E.E., M.K.;
-   Validation, I.E.M, A.E.E., A.K., M.K.;
-   Formal analysis, M.K., A.E.E., A.K.;
-   Data collection, M.K., A.E.E.;
-   Writing - original draft preparation, M.K.;
-   Writing - review and editing, I.E.M, A.E.E., A.K.;
-   Visualization, M.K.;
-   Supervision, I.E.M., A.E.E.;
-   Project administration, I.E.M.;
-   Funding acquisition, I.E.M.

All authors have read and agreed to the final version of the manuscript.

## CONSENT TO PUBLISH

All authors have given consent to publish.

## COMPETING INTERESTS

The authors declare no competing interests.

## DATA AVAILABILITY

## SUPPLEMENTARY INFORMATION

```{r VarSum}
#| label: tbl-varsum
#| tbl-cap: "Statistics summary table of selected quantitative variables from the fjord, landscape, grazing and fine-scale environment explanatory sets."
#| message: false
#| warning: false

# Packages
library(gt)
library(tidyverse)

# Join all variables
allvar <- purrr::reduce(list(subset(fjordsys, select = -c(JanTemp)), landscape, subset(grazing, select = -c(Sheep, Cattle)), subset(locenvi, select = -c(GWC, LOI, Humus, Nitrog))), dplyr::full_join)

# Summary statistics
varsum <- allvar |> 
  summarise(across(where(is.numeric), .fns = 
                     list(Mean = mean,
                          SD = sd,
                          Min = min,
                          Max = max
                     ))) |> 
  pivot_longer(everything(), names_sep = "_", names_to = c( "Variable", ".value")) 

# New set variable
varsum <- varsum |> 
  mutate(Set = ifelse(Variable == "Elev" |
                        Variable == "AnnPreci" |
                        Variable == "JulTemp" | 
                        Variable == "SeaDist", "Fjord", ifelse(
                          Variable == "Cultivated" |
                          Variable == "Forest" | 
                          Variable == "Infield" | 
                          Variable == "Outfield" |
                          Variable == "Wetland", "Landscape", ifelse(
                            Variable == "FieldSize" |
                            Variable == "FarmInfield" |
                            Variable == "StockDens", "Grazing", "Finescale"
    )
  )))

varsum |>
 # gt(groupname_col = "explanatory", rowname_col = "response") |>
 # tab_options(row_group.as_column = TRUE) |>
 gt(groupname_col = "Set", rowname_col = "Variable") |>
  # cols_label(
  #   Variance = "variance",
  #   PrF = "p-value",
  #   Df = "df",
  #   adj.r.squared = "adjRsq"
  # ) |>
  fmt_number(
    decimals = 1,
  ) |>
  fmt_number(
    decimals = 0,
    sep_mark = " ",
    rows = c("Elev", "AnnPreci", "SeaDist", "Aspect")
  ) |> 
  fmt_number(
    decimals = 2,
    rows = c("StockDens")
  ) |> 
  sub_missing(missing_text = "-") |>
  opt_stylize(style = 6, color = 'gray')

```


```{r RDAEnv}
#| label: tbl-rdaenv
#| tbl-cap: "Summary of statistical test for redundancy analyses between environmental variables."
#| message: false
#| warning: false

library(gt)
library(tidyverse)

rdaenv <- subset(rdastat_envi, select = -c(r.squared, type, eigenval))
rdaenv |> 
  gt(groupname_col = "model", rowname_col = "dim") |> 
  cols_move_to_start(
    columns = c("adj.r.squared")
  ) |>
  cols_move(
    columns = Df,
    after = F
  ) |>
  cols_label(
    Variance = "variance",
    PrF = "p-value",
    Df = "df",
    adj.r.squared = "adjRsq"
  ) |>
  fmt_number(
    columns = c("Variance", "F", "adj.r.squared"),
    decimals = 2
  ) |>
  fmt_number(
    columns = PrF,
    decimals = 4
  ) |>
  sub_missing(missing_text = "-") |> 
  opt_stylize(style = 6, color = 'gray')

```

```{r RDAGrass}
#| label: tbl-rdagrass
#| tbl-cap: "Summary of statistical test for redundancy analyses between grass assemblages and environmental variables."
#| message: false
#| warning: false

library(gt)
library(tidyverse)

rdagrass <- subset(rdastat_grass, select = -c(r.squared, type, eigenval))
rdagrass |> 
  gt(groupname_col = "model", rowname_col = "dim") |> 
  cols_move_to_start(
    columns = c("adj.r.squared")
  ) |>
  cols_move(
    columns = Df,
    after = F
  ) |>
  cols_label(
    Variance = "variance",
    PrF = "p-value",
    Df = "df",
    adj.r.squared = "adjRsq"
  ) |>
  fmt_number(
    columns = c("Variance", "F", "adj.r.squared"),
    decimals = 2
  ) |>
  fmt_number(
    columns = PrF,
    decimals = 4
  ) |>
  sub_missing(missing_text = "-") |> 
  opt_stylize(style = 6, color = 'gray')

```

```{r RDAForb}
#| label: tbl-rdaforb
#| tbl-cap: "Summary of statistical test for redundancy analyses between forb assemblage and environmental variables."
#| message: false
#| warning: false

library(gt)
library(tidyverse)

rdaforb <- subset(rdastat_forb, select = -c(r.squared, type, eigenval))
rdaforb |> 
  gt(groupname_col = "model", rowname_col = "dim") |> 
  cols_move_to_start(
    columns = c("adj.r.squared")
  ) |>
  cols_move(
    columns = Df,
    after = F
  ) |>
  cols_label(
    Variance = "variance",
    PrF = "p-value",
    Df = "df",
    adj.r.squared = "adjRsq"
  ) |>
  fmt_number(
    columns = c("Variance", "F", "adj.r.squared"),
    decimals = 2
  ) |>
  fmt_number(
    columns = PrF,
    decimals = 4
  ) |>
  sub_missing(missing_text = "-") |> 
  opt_stylize(style = 6, color = 'gray')

```

```{r RDABeetle}
#| label: tbl-rdabeetle
#| tbl-cap: "Summary of statistical test for redundancy analyses between beetle assemblage and environmental variables."
#| message: false
#| warning: false

library(gt)
library(tidyverse)

rdabeetle <- subset(rdastat_beetle, select = -c(r.squared, type, eigenval))
rdabeetle |> 
  gt(groupname_col = "model", rowname_col = "dim") |> 
  cols_move_to_start(
    columns = c("adj.r.squared")
  ) |>
  cols_move(
    columns = Df,
    after = F
  ) |>
  cols_label(
    Variance = "variance",
    PrF = "p-value",
    Df = "df",
    adj.r.squared = "adjRsq"
  ) |>
  fmt_number(
    columns = c("Variance", "F", "adj.r.squared"),
    decimals = 2
  ) |>
  fmt_number(
    columns = PrF,
    decimals = 4
  ) |>
  sub_missing(missing_text = "-") |> 
  opt_stylize(style = 6, color = 'gray')

```

<!-- ```{r FarmerdManagement} -->

<!-- #| label: tbl-farmermanagement -->

<!-- #| tbl-cap: "Field management applied by the farmer for the last 10 years." -->

<!-- #| message: true -->

<!-- #| warning: true -->

<!-- library(gt) -->

<!-- farmer_management <- subset(landuse_grass, select = c(SiteID, FM_herbicide, FM_treecutting, FM_burning, FM_mowing, FM_mulching, FM_manure, FM_artificialfert, FM_shellsandlime, Lastknownre_sowing, Lastknownplowing, Lastknowndrainage)) -->

<!-- farmer_management |>  -->

<!--   gt() |>  -->

<!--   cols_label(SiteID = "Site", -->

<!--              FM_herbicide = "Herbicide", -->

<!--              FM_treecutting = "Tree cutting", -->

<!--              FM_burning = "Burning", -->

<!--              FM_mowing = "Mowing", -->

<!--              FM_mulching = "Mulching", -->

<!--              FM_manure = "Manure", -->

<!--              FM_artificialfert = "Artificial fertilizer", -->

<!--              FM_shellsandlime = "Shell/sand/lime", -->

<!--              Lastknownre_sowing = "Last known re-sowing", -->

<!--              Lastknownplowing = "Last known plowing", -->

<!--              Lastknowndrainage = "Last known drainage") |>  -->

<!--   tab_spanner(label = "Fertilization", columns = c(FM_manure, FM_artificialfert, FM_shellsandlime)) -->

<!-- ``` -->

<!-- ```{r VarSummary} -->

<!-- #| label: tbl-varsummary -->

<!-- #| tbl-cap: "Summary of explanatory variables for canonical correlation analysis of grass, forb and beetle dominant community data. Symbols for rejection are: ¤ non-normal distribution; ◦ colinearity with another variable; * linearity assumption rejected." -->

<!-- #| message: false -->

<!-- #| warning: false -->

<!-- library(readxl) -->

<!-- library(gt) -->

<!-- library(tidyverse) -->

<!-- varsummary <- read_xlsx("illustrations/FjordModelFactors.xlsx") -->

<!-- # summary <- subset(papersummary, select = -c(Group)) -->

<!-- varsummary <- varsummary |> -->

<!--   group_by(Variable_set) |> -->

<!--   pivot_wider(names_from = Assumption, values_from = Confirmation) -->

<!-- varsummary |> -->

<!--   gt(groupname_col = "Variable_set") |> -->

<!--   cols_label( -->

<!--     Variable = "" -->

<!--     ) |> -->

<!--   opt_stylize(style = 6, color = 'gray') -->

<!-- ``` -->
